Estimating Gaze Duration Error Distributions from Eye Tracking Data

### 2022-06-20 paper/paper.tex
Adding Bayesian posterior estimation

Eye tracking is an important technology with a wide variety of applications. It permits evaluation
of software user interfaces\cite{Harezlak2015}, the development of new forms of software interaction, as well as a potential method of biometric identification\cite{Kasprowski2018}. 
Eye tracking has become a core tool for empirical investigations 
into human behaviour and is widely used as a method of measuring explicit attention to visual stimuli. 
It has been applied to study psychological phenomena ranging from cognition\cite{Brunye2019}
to mental health\cite{Duque2014,Rudich-Strassler2022}, and is now routinely used to evaluate 
advertising \cite{Hervet2011}. Eye tracking technology has allowed the marketing researchers 
to study many factors that contribute to effective advertising, 
including brand recall \cite{Wedel2000},
the capture and tranfer of attention \cite{Pieters2004}, 
the impact of images of faces \cite{Djamasbi2010},
the attention effects of animation \cite{Hamborg2012},
and the relationship with social media posts \cite{Barreto2013}.

@@

When fixation models are used sequentially for the purpose of estimating gaze duration
(as is done in advertising media studies), then there is the potential for the error
to either cancel out (reducing error in gaze duration estimation) or to compound
(increasing error in gaze duration). Which of these outcomes occurs will depend on the
specific of the model and the conditions under which it is being used.
Influencing factors include the ratio of true positives to true negatives in the data,
and the error profile of the machine learning fixation coordinate model.

The errors of eye tracking data technology can be decomposed into a range of
independent sources that researchers need to consider \cite{Holmqvist2012}.
Unless they are properly address these sources of error can manifest themselves
as systematic biases across undesirable dimensions such as the age\cite{Dalrymple2018}
or ethnicity of subjects\cite{Blignaut2013}.

The use of eye fixation models to estimate gaze duration (or dwell time) in Area of
Interest studies (AOI) was discussed by Holmqvist et al. \cite{Holmqvist2012}. The
authors simulated the impact of gaze fixation error by adding noise accoridng to
manufacturer specifications to data sets of low margin areas of interest. We extend 
this idea to calculate probabilistic bounds on the error in gaze duration using
calibration data as a source of noise distribution that is specific to both the
study participants and the device/environment of the study. The result is an open
source application that may be used by a wide variety of reserachers to provide
error bounds on any gaze duration measurements.

@@

In the first stage of the methodology we provide an estimation of the distribution of 
measurements for a given true gaze duration. These distributions are produced at multiple locations
across the screen, so that we have a set of distributions, that captures the expected
variation in measurement depending on the length of and locus of true fixation.

We want to estimate a the probability distribution over measured durations $\hat{d}$ given a
known duration $d$, the session length $s$ and the ad location $l$. 
We express this distribution as $P(\hat{d}|d,l,s)$. The
procedure for making this estimation involves iteratively generating random samples
of gaze fixation paths $\mathcal{F}$ consistent with the parameters $d$, $s$ and $l$. 
For each point $f$ in $\mathcal{F}$ we determine a measurement point $\hat{f}$ 
by drawing samples from the eye tracking validation set. 
These points determine the $\delta_x$ and $\delta_y$ error in the measurement
of gaze fixation point $f$. We draw these coordinates from the validation file such the
the probability of the point being chosen is propotional to the distance of its target
$x$ and $y$ from the coordinates of the point $f$ in the generated path. The complete algorithm
for estimating $P(\hat{d}|d,l,s)$ is shown in Algorithm \ref{alg:p_of_dhat}

\begin{algorithm}
\caption{Estimation of $P(\hat{d}|d,l,s)$}\label{alg:p_of_dhat}
\begin{algorithmic}
  \Require $\hat{d},l,s,C$
  \Ensure $P(\hat{d}|d,l,s)$
  \State $N \gets 2000$
  \State $D \gets floor(s \times 10)$ \Comment{Durations sampled at 100ms intervals}
  \State $P \gets Array(D+1,D+1)$ \Comment{Result is a 2 dimensional array}
  \If{$\hat{d} > s$}
    \State error
  \EndIf
  \For{$d \in 0 to D$} \Comment{Iterate over all D+1 possible durations}
    \For{$n \in 1 to N$} \Comment{Collect N samples}
      \State $p \gets generatePath(d,l,s)$
      \State $d_e \gets generateMeasurement(p,C)$  
      \State $P[d,d_e] += 1/2000$  
    \EndFor
  \EndFor
\end{algorithmic}
\end{algorithm}

The bulk of the work is done by two functions inside the simulation loop. The first function
$generatePath(d,l,s)$ takes a true gaze duration, position of ad and viewing session length and
generates a path of fixation points compatible with those parameters. The second function 
$generateMeasurement(p,C)$, takes the path and the eye tracking calibration data and produces
an sample of measured attention time. Repeated application of this process over different 
values of $d$ allows us to populate a two dimensional array where the first dimension is the
true duration and the second the estimated. Note: that our probability distribution is discrete
because we restrict ourselves to sampling every 100ms point between zero and the session length.

>>>>

Eye tracking is an important technology with a wide variety of applications. It permits evaluation
of software user interfaces\cite{Harezlak2015}, the development of new forms of software interaction, as well as a potential method of biometric identification\cite{Kasprowski2018}. 
Eye tracking has become a core tool for empirical investigations 
into human behaviour and is widely used as a method of measuring explicit attention to visual stimuli. 
It has been applied to study psychological phenomena ranging from cognition\cite{Brunye2019}
to mental health\cite{Duque2014,Rudich-Strassler2022}, and is now routinely used to evaluate 
advertising \cite{Hervet2011}. Eye tracking technology has allowed marketing researchers 
to study many factors that contribute to effective advertising, 
including brand recall \cite{Wedel2000},
the capture and tranfer of attention \cite{Pieters2004}, 
the impact of images of faces \cite{Djamasbi2010},
the attention effects of animation \cite{Hamborg2012},
and the relationship with social media posts \cite{Barreto2013}.

@@

When fixation models are used sequentially for the purpose of estimating gaze duration
(as is commonly done in point of salience or advertising media studies), 
then there is the potential for the error
to either cancel out (reducing error in gaze duration estimation) or to compound
(increasing error in gaze duration). Which of these outcomes occurs will depend on the
specifics of the model and the conditions under which it is being used.
Influencing factors include the ratio of measured gaze duration to the total viewing time
and the error profile of the machine learning fixation coordinate model.

The errors of eye tracking data technology can be decomposed into a range of
independent sources that researchers need to consider \cite{Holmqvist2012}.
Unless they are properly addressed these sources of error can manifest themselves
as systematic biases across undesirable dimensions such as the age\cite{Dalrymple2018}
or ethnicity of subjects\cite{Blignaut2013}.

The use of eye fixation models to estimate gaze duration (or dwell time) in Area of
Interest studies (AOI) was discussed by Holmqvist et al. \cite{Holmqvist2012}. The
authors simulated the impact of gaze fixation error by adding noise according to
manufacturer specifications to data sets of low margin areas of interest. We extend 
this idea to calculate probabilistic bounds on the error in gaze duration using
calibration data as a source of noise distribution that is specific to both the
study participants and the device/environment of the study. The result is an open
source application that may be used by a wide variety of reserachers to provide
error bounds on any gaze duration measurements.

@@

In the first stage of the methodology we provide an estimate of the distribution of measured
durations for all possible true gaze durations. These distributions are produced by generating
random fixation paths across the available screen dimensions, and adding noise that is 
drawn from the calibration data to make it consistent with the observed properties of the gaze model.
This produces a set of distributions that capture the expected
variation in measurement depending on the length of true fixation.

Formally stated this is the probability distribution over measured durations $\hat{d}$ given a
known duration $d$, the session length $s$ and the gaze target location $l$. 
We express this distribution as $P(\hat{d}|d,l,s)$. The
procedure for making this estimation involves iteratively generating random samples
of gaze fixation paths $\mathcal{F}$ consistent with the parameters $d$, $s$ and $l$. 
For each point $f$ in $\mathcal{F}$ we determine a measurement point $\hat{f}$ 
by drawing samples from the eye tracking validation set. 
These points determine the $\delta_x$ and $\delta_y$ error in the measurement
of gaze fixation point $f$. We draw these coordinates from the Calibration file such the
the probability of the point being chosen is propotional to the distance of its target
$x$ and $y$ from the coordinates of the point $f$ in the generated path. The complete algorithm
for estimating $P(\hat{d}|d,l,s)$ is shown in Algorithm \ref{alg:p_of_dhat}

\begin{algorithm}
\caption{Estimation of $P(\hat{d}|d,l,s)$}\label{alg:p_of_dhat}
\begin{algorithmic}
  \Require $l,s,C$
  \Ensure $P(\hat{d}|d,l,s)$
  \State $N \gets samples$              \Comment{Configure simulation samples per duration}
  \State $D \gets floor(s / increment)$ \Comment{Durations are sampled at discrete intervals}
  \State $P \gets Array(D+1,D+1)$       \Comment{The distribution P is a 2 dimensional array}
  \For{$d \in 0 to D$}                  \Comment{Iterate over all possible true gaze durations}
    \For{$n \in 0 to N$}                \Comment{Iterature to collect the N samples for d}
      \State $p \gets generatePath(d,l,s)$
      \State $d_e \gets generateMeasurement(p,C)$  
      \State $P[d,d_e] += 1/N$          \Comment{Probability increment for a measuremnt of $d_e$}
    \EndFor
  \EndFor
  \State return P
\end{algorithmic}
\end{algorithm}

The bulk of the work is done by two functions inside the simulation loop. The first function
$generatePath(d,l,s)$ takes a true gaze duration (for that simulation iteration) the location
of the gaze target and the viewing session length. It then
generates a random path of fixation points that is compatible with those parameters.
The second function 
$generateMeasurement(p,l,C)$, takes the path, target location and the eye tracking calibration 
data and produces a sample of measured attention time. It does this by using the calibration data to sample
fixation errors and apply them to the path data. The noisy path data is then used to determine
the estimated gaze duration by looking at the number of fixation points that intersect with the target
location.
Repeated application of this process for a given value of $d$ produces multiple samples of $\hat{d}$
for that true gaze duration. Repeating the process over different 
values of $d$ allows us to populate a two dimensional array where the first dimension is the
true duration and the second the estimated duration. Note that our probability distribution is discrete
because we restrict ourselves to sampling every increment point between zero and the session length.
